---
title: "Mapping fishing effort beyond AIS"
subtitle: "Combining vessel broadcasts with Sentinel-1 satellite radar"
author: "Théophile L. Mouton"
date: today
format:
  html:
    embed-resources: true
    css: Quarto_template/custom.css
    include-before-body: Quarto_template/header.html
    include-after-body: Quarto_template/footer.html
    theme: none
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    fig-align: center
execute:
  echo: true
  message: false
  warning: false
---

```{r libs}
#| echo: false
#| output: false
library(dplyr)
library(ggplot2)
library(maps)
library(scales)
library(patchwork)
library(knitr)
```

## The problem

Public monitoring of industrial fishing leans heavily on AIS, the transponder
system vessels use to broadcast their position. Coverage is uneven: not every
vessel carries it, not every vessel keeps it switched on, and satellite reception
varies by region. Sentinel-1 radar sees vessels regardless of what they broadcast,
but a radar detection tells you only that something was there at that moment, not
how long it fished.

This document builds one figure from three layers: what vessels report, what
satellites see, and a model that estimates fishing hours in the cells where radar
finds vessels and AIS is silent.

Both sources come from Global Fishing Watch, aggregated to a 0.1 degree global
grid over 2017 to 2020.

## Loading the data

Panels A and B come straight from the pre-aggregated grid, so the raw AIS and SAR
records are never opened.

```{r load}
#| output: false
load(here::here("R", "Data", "combined_data_O1deg.Rdata"))

ais_col <- "total_fishing_hours"
sar_col <- "total_presence_score"

d <- as.data.frame(combined_data_01deg)

panel_a <- d %>%
  filter(has_AIS, .data[[ais_col]] > 0) %>%
  transmute(lon_std, lat_std, value = .data[[ais_col]])

panel_b <- d %>%
  filter(has_SAR, .data[[sar_col]] > 0) %>%
  transmute(lon_std, lat_std, value = .data[[sar_col]])
```

## Predicting fishing hours from radar

The model is a random forest trained on cells where both signals exist, predicting
log fishing hours from SAR detections per satellite overpass, coordinates,
distance to shore, distance to port and bathymetry.

Detections are normalised by overpass frequency, and that choice matters more than
it might appear. Cells overflown more often accumulate more detections regardless
of how much fishing happens there, so a model trained on raw counts partly learns
the satellite's orbit rather than the fishery. Removing that artefact costs some
apparent accuracy and buys a model that measures the right thing.

```{r prepare}
#| output: false
cache <- here::here("R", "Data", "figure_cache.Rdata")

if (file.exists(cache)) {
  load(cache)
} else {
  library(data.table)
  library(raster)
  library(randomForest)

  load(here::here("R", "Data", "SAR_overpasses_per_cell.Rdata"))

  raster_df <- as.data.frame(
    stack(raster(here::here("R", "Data", "distance-from-shore-0.1deg-adjusted.tif")),
          raster(here::here("R", "Data", "distance-from-port-0.1deg-adjusted.tif")),
          raster(here::here("R", "Data", "bathymetry-0.1deg-adjusted.tif"))),
    xy = TRUE)
  names(raster_df) <- c("x", "y", "dist_shore", "dist_ports", "bathy")
  raster_df <- na.omit(raster_df)
  setDT(raster_df)
  raster_df[, `:=`(lon_std = round(x, 1), lat_std = round(y, 1))]
  raster_df <- raster_df[bathy < 0]
  raster_df <- unique(raster_df, by = c("lon_std", "lat_std"))

  dt <- merge(
    as.data.table(d),
    as.data.table(overpasses_per_cell)[, .(lon_std, lat_std, detections_per_overpass)],
    by = c("lon_std", "lat_std"), all.x = TRUE
  )
  dt <- unique(dt, by = c("lon_std", "lat_std"))

  with_rasters <- as.data.frame(
    raster_df[dt, on = .(lon_std, lat_std), nomatch = 0]
  )

  # Cells seen by radar but not broadcasting: the prediction targets
  sar_only <- with_rasters %>%
    filter(has_SAR, !has_AIS) %>%
    dplyr::select(detections_per_overpass, lon_std, lat_std,
                  dist_shore, dist_ports, bathy) %>%
    na.omit()

  rf <- readRDS(here::here("R", "Data", "rf_norm_fishing_log.rds"))
  sar_only$value <- 10^predict(rf, newdata = sar_only) - 1
  panel_c <- sar_only %>% dplyr::select(lon_std, lat_std, value)

  # Cells with both signals: the validation set
  validation <- with_rasters %>%
    filter(has_AIS, has_SAR) %>%
    mutate(log_total_fishing_hours = log10(total_fishing_hours + 1),
           log_detections_per_overpass = log10(detections_per_overpass + 1))

  evaluate_model <- function(model, data, log_target = FALSE) {
    p <- predict(model, newdata = data)
    if (log_target) p <- 10^p - 1
    actual <- if (log_target) 10^data$log_total_fishing_hours - 1 else data$total_fishing_hours
    res <- actual - p
    n <- length(actual)
    k <- length(attr(model$terms, "term.labels"))
    r2 <- model$rsq[length(model$rsq)]
    c(mae    = mean(abs(res), na.rm = TRUE),
      rmse   = sqrt(mean(res^2, na.rm = TRUE)),
      mape   = mean(abs(res / actual) * 100, na.rm = TRUE),
      medae  = median(abs(res), na.rm = TRUE),
      r2     = r2,
      adj_r2 = 1 - ((1 - r2) * (n - 1) / (n - k - 1)),
      m_res  = mean(res, na.rm = TRUE),
      sd_res = sd(res, na.rm = TRUE))
  }

  m_none      <- readRDS(here::here("R", "Data", "rf_norm_no_transform.rds"))
  m_fish_log  <- rf
  m_pred_log  <- readRDS(here::here("R", "Data", "rf_norm_predictor_log.rds"))
  m_both_log  <- readRDS(here::here("R", "Data", "rf_norm_both_log.rds"))

  results_comparison <- data.frame(
    Metric = c("Mean absolute error", "Root mean squared error",
               "Mean absolute percentage error", "Median absolute error",
               "R-squared", "Adjusted R-squared",
               "Mean of residuals", "SD of residuals"),
    `No transform`      = round(evaluate_model(m_none, validation), 2),
    `Fishing hours log` = round(evaluate_model(m_fish_log, validation, TRUE), 2),
    `Predictor log`     = round(evaluate_model(m_pred_log, validation), 2),
    `Both log`          = round(evaluate_model(m_both_log, validation, TRUE), 2),
    check.names = FALSE, row.names = NULL
  )

  save(panel_c, results_comparison, file = cache)
}
```

### Comparing transformations

Four variants were compared, differing only in whether the response, the predictor
or both were log transformed.

```{r table}
kable(results_comparison, format = "html", digits = 2,
      caption = "Model performance on cells where AIS and SAR both report")
```

Log transforming the response is what matters. The two variants that do so reach
an R-squared of 0.77 and cut the mean absolute percentage error from well over a
thousand percent to roughly eighty-six, while the median absolute error falls from
around thirty-three hours to about twelve. Log transforming the predictor alone
changes almost nothing.

The fishing hours log model is the one used here. It edges out the both-log
variant on error metrics and is easier to interpret, since detections per overpass
stay on their original scale.

## The figure

```{r panels}
#| output: false
world <- map_data("world")

make_panel <- function(dat, legend_title, breaks) {
  ggplot() +
    geom_map(data = world, map = world,
             aes(long, lat, map_id = region),
             color = NA, fill = "grey86", linewidth = 0) +
    geom_tile(data = dat, aes(lon_std, lat_std, fill = value)) +
    scale_fill_viridis_c(
      option = "inferno", direction = -1, trans = "log1p",
      name = legend_title, breaks = breaks, labels = comma,
      guide = guide_colorbar(barwidth = 14, barheight = 0.4,
                             title.position = "top", title.hjust = 0.5)
    ) +
    coord_fixed(1.15, xlim = c(-180, 180), ylim = c(-70, 82), expand = FALSE) +
    theme_void(base_size = 13) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(size = 11, margin = ggplot2::margin(b = 4)),
      legend.text = element_text(size = 9),
      plot.title = element_text(face = "bold", size = 14,
                                margin = ggplot2::margin(b = 2)),
      plot.subtitle = element_text(size = 11, colour = "grey30",
                                   margin = ggplot2::margin(b = 4)),
      plot.margin = ggplot2::margin(4, 8, 4, 8)
    )
}

hour_breaks <- c(0, 10, 1000, 100000)

pa <- make_panel(panel_a, "Fishing hours, 2017 to 2020", hour_breaks) +
  labs(title = "What vessels report",
       subtitle = "AIS broadcast fishing effort")

pb <- make_panel(panel_b, "Summed detection score, 2017 to 2020",
                 c(0, 10, 100, 1000)) +
  labs(title = "What satellites see",
       subtitle = "Sentinel-1 radar detections of fishing vessels")

pc <- make_panel(bind_rows(panel_a, panel_c),
                 "Fishing hours, 2017 to 2020", hour_breaks) +
  labs(title = "The complete picture",
       subtitle = "Reported hours plus model estimates for radar-detected cells")

fig <- pa / pb / pc
```

Panels A and C share a colour scale, so any difference between them is a
difference in effort rather than an artefact of rescaling.

```{r figure}
#| fig-width: 8
#| fig-height: 13
#| out-width: "62%"
#| echo: false
fig
```

```{r save}
#| echo: false
#| output: false
dir.create(here::here("R", "Outputs"), showWarnings = FALSE, recursive = TRUE)
ggsave(here::here("R", "Outputs", "fishing_effort_panels.png"),
       plot = fig, width = 8, height = 13, dpi = 150, bg = "white")
```

::: {.callout-note}
Data from Global Fishing Watch: AIS fishing effort (Kroodsma et al. 2018,
*Science*) and Sentinel-1 vessel detections (Paolo et al. 2024, *Nature*).
Figure supports a manuscript in preparation.
:::
