---
title: "Mapping industrial fishing effort beyond AIS"
subtitle: "Combining vessel broadcasts with Sentinel-1 satellite radar"
author: "Théophile L. Mouton"
date: today
format:
  html:
    output-file: index.html
    embed-resources: true
    css: Quarto_template/custom.css
    include-before-body: Quarto_template/header.html
    include-after-body: Quarto_template/footer.html
    theme: none
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    fig-align: center
execute:
  echo: true
  message: false
  warning: false
---

```{r libs}
#| echo: false
#| output: false
library(dplyr)
library(ggplot2)
library(maps)
library(scales)
library(patchwork)
library(knitr)
```

## The problem

Public monitoring of industrial fishing leans heavily on AIS, the transponder
system vessels use to broadcast their position. Coverage is uneven: not every
vessel carries it, not every vessel keeps it switched on, and satellite reception
varies by region. Sentinel-1 radar sees vessels regardless of what they broadcast,
and deep learning classifies each detection as fishing or non-fishing, but a
detection is a snapshot. It records presence at one moment, not time spent
fishing.

The goal here is to convert presence into effort: to estimate fishing hours in
the cells where radar finds vessels and AIS is silent, and so extend a global
effort layer beyond what broadcast data alone can cover.

Both sources come from Global Fishing Watch, aggregated to a 0.1 degree global
grid over 2017 to 2020.

## Loading the data

The grid is pre-aggregated, so the raw AIS and SAR records are never opened.

```{r load}
#| output: false
load(here::here("R", "Data", "combined_data_O1deg.Rdata"))

ais_col <- "total_fishing_hours"
sar_col <- "total_presence_score"

d <- as.data.frame(combined_data_01deg)

panel_a <- d %>%
  filter(has_AIS, .data[[ais_col]] > 0) %>%
  transmute(lon_std, lat_std, value = .data[[ais_col]])

panel_b <- d %>%
  filter(has_SAR, .data[[sar_col]] > 0) %>%
  transmute(lon_std, lat_std, value = .data[[sar_col]])
```

### How the two signals overlap

```{r coverage}
data.frame(
  Category = c("AIS only",
               "SAR only",
               "Both signals",
               "Total cells with any signal"),
  Cells = c(
    sum(d$has_AIS & !d$has_SAR),
    sum(!d$has_AIS & d$has_SAR),
    sum(d$has_AIS & d$has_SAR),
    sum(d$has_AIS | d$has_SAR)
  )
) %>%
  kable(format = "html", format.args = list(big.mark = " "),
        caption = "Grid cells at 0.1 degree resolution, 2017 to 2020")
```

Cells carrying both signals are what the model learns from. Cells with radar
detections and no AIS are what it is applied to.

One caveat worth stating plainly: Sentinel-1 imagery used by Paolo et al. covers
roughly 15 percent of the ocean, concentrated in coastal waters where most
industrial activity occurs. Empty ocean in the SAR panel below means no imagery,
not an absence of fishing.

## Predicting fishing hours from radar

The model is a random forest trained on cells where both signals exist,
predicting log fishing hours from SAR detections per satellite overpass,
coordinates, distance to shore, distance to port and bathymetry.

Detections are normalised by overpass frequency, and that choice matters more
than it might appear. Cells overflown more often accumulate more detections
regardless of how much fishing happens there, so a model trained on raw counts
partly learns the satellite's orbit rather than the fishery. Paolo et al. report
activity the same way, as detections per overpass, for the same reason.

An earlier version of this model used raw counts and reached an R-squared of
0.82. Removing the confound costs some apparent accuracy and buys a model that
measures the right thing. The comparison below covers only the normalised
variants.

```{r prepare}
#| output: false
cache <- here::here("R", "Data", "figure_cache.Rdata")

if (file.exists(cache)) {
  load(cache)
} else {
  library(data.table)
  library(raster)
  library(randomForest)

  load(here::here("R", "Data", "SAR_overpasses_per_cell.Rdata"))

  raster_df <- as.data.frame(
    stack(raster(here::here("R", "Data", "distance-from-shore-0.1deg-adjusted.tif")),
          raster(here::here("R", "Data", "distance-from-port-0.1deg-adjusted.tif")),
          raster(here::here("R", "Data", "bathymetry-0.1deg-adjusted.tif"))),
    xy = TRUE)
  names(raster_df) <- c("x", "y", "dist_shore", "dist_ports", "bathy")
  raster_df <- na.omit(raster_df)
  setDT(raster_df)
  raster_df[, `:=`(lon_std = round(x, 1), lat_std = round(y, 1))]
  raster_df <- raster_df[bathy < 0]
  raster_df <- unique(raster_df, by = c("lon_std", "lat_std"))

  dt <- merge(
    as.data.table(d),
    as.data.table(overpasses_per_cell)[, .(lon_std, lat_std, detections_per_overpass)],
    by = c("lon_std", "lat_std"), all.x = TRUE
  )
  dt <- unique(dt, by = c("lon_std", "lat_std"))

  with_rasters <- as.data.frame(
    raster_df[dt, on = .(lon_std, lat_std), nomatch = 0]
  )

  # Cells seen by radar but not broadcasting: the prediction targets
  sar_only <- with_rasters %>%
    filter(has_SAR, !has_AIS) %>%
    dplyr::select(detections_per_overpass, lon_std, lat_std,
                  dist_shore, dist_ports, bathy) %>%
    na.omit()

  rf <- readRDS(here::here("R", "Data", "rf_norm_fishing_log.rds"))
  sar_only$value <- 10^predict(rf, newdata = sar_only) - 1
  panel_c <- sar_only %>% dplyr::select(lon_std, lat_std, value)

  # Cells with both signals: the validation set
  validation <- with_rasters %>%
    filter(has_AIS, has_SAR) %>%
    mutate(log_total_fishing_hours = log10(total_fishing_hours + 1),
           log_detections_per_overpass = log10(detections_per_overpass + 1))

  n_train <- nrow(validation)

  evaluate_model <- function(model, data, log_target = FALSE) {
    p <- predict(model, newdata = data)
    if (log_target) p <- 10^p - 1
    actual <- if (log_target) 10^data$log_total_fishing_hours - 1 else data$total_fishing_hours
    res <- actual - p
    n <- length(actual)
    k <- length(attr(model$terms, "term.labels"))
    r2 <- model$rsq[length(model$rsq)]
    c(mae    = mean(abs(res), na.rm = TRUE),
      rmse   = sqrt(mean(res^2, na.rm = TRUE)),
      mape   = mean(abs(res / actual) * 100, na.rm = TRUE),
      medae  = median(abs(res), na.rm = TRUE),
      r2     = r2,
      adj_r2 = 1 - ((1 - r2) * (n - 1) / (n - k - 1)),
      m_res  = mean(res, na.rm = TRUE),
      sd_res = sd(res, na.rm = TRUE))
  }

  m_none      <- readRDS(here::here("R", "Data", "rf_norm_no_transform.rds"))
  m_fish_log  <- rf
  m_pred_log  <- readRDS(here::here("R", "Data", "rf_norm_predictor_log.rds"))
  m_both_log  <- readRDS(here::here("R", "Data", "rf_norm_both_log.rds"))

  results_comparison <- data.frame(
    Metric = c("Mean absolute error", "Root mean squared error",
               "Mean absolute percentage error", "Median absolute error",
               "R-squared", "Adjusted R-squared",
               "Mean of residuals", "SD of residuals"),
    `No transform`      = round(evaluate_model(m_none, validation), 2),
    `Fishing hours log` = round(evaluate_model(m_fish_log, validation, TRUE), 2),
    `Predictor log`     = round(evaluate_model(m_pred_log, validation), 2),
    `Both log`          = round(evaluate_model(m_both_log, validation, TRUE), 2),
    check.names = FALSE, row.names = NULL
  )

  save(panel_c, results_comparison, n_train, file = cache)
}

# Older caches predate n_train; fall back to the pre-join count
if (!exists("n_train")) n_train <- sum(d$has_AIS & d$has_SAR)
```

### Comparing transformations

Four variants were compared, differing only in whether the response, the
predictor or both were log transformed.

```{r table}
kable(results_comparison, format = "html", digits = 2,
      caption = "Model performance on cells where AIS and SAR both report")
```

Log transforming the response is what matters. The two variants that do so reach
an R-squared of 0.77 and cut the mean absolute percentage error from well over a
thousand percent to roughly eighty-six, while the median absolute error falls
from around thirty-three hours to about twelve. Log transforming the predictor
alone changes almost nothing.

The fishing hours log model is the one used here. It edges out the both-log
variant on error metrics and is easier to interpret, since detections per
overpass stay on their original scale.

## What this recovers

```{r gain}
n_ais <- nrow(panel_a)
n_new <- nrow(panel_c)

data.frame(
  Measure = c("Cells used to train the model",
              "Cells with AIS-based effort",
              "Cells gained from radar predictions",
              "Increase over AIS coverage (%)"),
  Value = c(n_train, n_ais, n_new, round(100 * n_new / n_ais, 1))
) %>%
  kable(format = "html", format.args = list(big.mark = " "),
        caption = "Coverage gained, 0.1 degree cells")
```

The percentage is expressed against AIS coverage rather than against all ocean
cells, since cells outside the SAR footprint were never candidates for
prediction.

## The figure

```{r panels}
#| output: false
world <- map_data("world")

make_panel <- function(dat, legend_title, breaks) {
  ggplot() +
    geom_map(data = world, map = world,
             aes(long, lat, map_id = region),
             color = NA, fill = "grey86", linewidth = 0) +
    geom_tile(data = dat, aes(lon_std, lat_std, fill = value)) +
    scale_fill_viridis_c(
      option = "inferno", direction = -1, trans = "log1p",
      name = legend_title, breaks = breaks, labels = comma,
      guide = guide_colorbar(barwidth = 14, barheight = 0.4,
                             title.position = "top", title.hjust = 0.5)
    ) +
    coord_fixed(1.15, xlim = c(-180, 180), ylim = c(-70, 82), expand = FALSE) +
    theme_void(base_size = 13) +
    theme(
      legend.position = "bottom",
      legend.title = element_text(size = 11, margin = ggplot2::margin(b = 4)),
      legend.text = element_text(size = 9),
      plot.title = element_text(face = "bold", size = 14,
                                margin = ggplot2::margin(b = 2)),
      plot.subtitle = element_text(size = 11, colour = "grey30",
                                   margin = ggplot2::margin(b = 4)),
      plot.margin = ggplot2::margin(4, 8, 4, 8)
    )
}

hour_breaks <- c(0, 10, 1000, 100000)

pa <- make_panel(panel_a, "Fishing hours, 2017 to 2020", hour_breaks) +
  labs(title = "What vessels report",
       subtitle = "AIS broadcast fishing effort")

pb <- make_panel(panel_b, "Summed detection score, 2017 to 2020",
                 c(0, 10, 100, 1000)) +
  labs(title = "What satellites see",
       subtitle = "Sentinel-1 radar detections of fishing vessels")

pc <- make_panel(bind_rows(panel_a, panel_c),
                 "Fishing hours, 2017 to 2020", hour_breaks) +
  labs(title = "The complete picture",
       subtitle = "Reported hours plus model estimates for radar-detected cells")

fig <- pa / pb / pc
```

Panels A and C share a colour scale, so any difference between them is a
difference in effort rather than an artefact of rescaling.

```{r figure}
#| fig-width: 8
#| fig-height: 13
#| out-width: "62%"
#| echo: false
fig
```

```{r save}
#| echo: false
#| output: false
dir.create(here::here("R", "Outputs"), showWarnings = FALSE, recursive = TRUE)
ggsave(here::here("R", "Outputs", "fishing_effort_panels.png"),
       plot = fig, width = 8, height = 13, dpi = 150, bg = "white")
```

::: {.callout-note}
Data from Global Fishing Watch: AIS fishing effort (Kroodsma et al. 2018,
*Science*) and Sentinel-1 vessel detections (Paolo et al. 2024, *Nature*).

This layer supports Mouton et al., *Global gaps and priorities for shark and ray
conservation: integrating threat, function, and evolutionary distinctiveness*,
preprint at [doi.org/10.1101/2025.11.28.691085](https://doi.org/10.1101/2025.11.28.691085).
:::
